---
title: "Numbers used in text of the manuscript"
author: "Mathew Kiang"
date: "`r Sys.Date()`"
output: 
  html_document:
    code_folding: 'hide'
    toc: true
    toc_float: true
---

```{r options, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(error = FALSE)
```

## Set up

```{r settings, results='hide'}
library(tidyverse)
library(here)
library(knitr)
library(kableExtra)
source(here("code", "utils.R"))
```

```{r data}
results_df <- readRDS(here("data", "summarized_results.RDS")) |> 
    categorize_race() |> 
    categorize_deaths()
results_age_df <- readRDS(here("data", "summarized_results_by_age.RDS")) |> 
    categorize_race() |> 
    categorize_deaths()
demographics <- readRDS(here("data", "death_demographics.RDS"))
```

## Demographics of deaths
```{r demographics}
TOTAL_DEATHS <- sum(demographics$n_cancer)
MEAN_AGE <- weighted.mean(demographics$age_years, demographics$n_cancer, na.rm = TRUE)
SD_AGE <- sd(rep(demographics$age_years, times = demographics$n_cancer), na.rm = TRUE)
MEDIAN_AGE <- median(rep(demographics$age_years, times = demographics$n_cancer), na.rm = TRUE)
IQRL_AGE <- quantile(rep(demographics$age_years, times = demographics$n_cancer), probs = .25, na.rm = TRUE)
IQRU_AGE <- quantile(rep(demographics$age_years, times = demographics$n_cancer), probs = .75, na.rm = TRUE)

by_sex <- demographics |> group_by(sex) |> summarize(n = sum(n_cancer), .groups = "drop") |> mutate(prop = n / sum(n))

by_raceeth <- demographics |> group_by(race_eth) |> summarize(n = sum(n_cancer), .groups = "drop") |> mutate(prop = n / sum(n))

by_sex
by_raceeth
```
Between 1999 and 2020, there were `r prettyNum(TOTAL_DEATHS, big.mark = ",")` deaths due to cancer with a mean (SD) age of `r round(MEAN_AGE, 1)` (`r round(SD_AGE, 1)`) years. Most deaths occurred among males (52.3%; n = 6,608,436) and White decedents (80.1%; n = 10,124,360) followed by Black (11.4%; n = 1,439,259) and Hispanic (5.7%; n = 717,080) decedents. 

Between 1999 and 2020, there were `r prettyNum(TOTAL_DEATHS, big.mark = ",")` deaths due to cancer with a median (IQR) age of `r round(MEDIAN_AGE)` (`r IQRL_AGE` to `r IQRU_AGE`) years. Most deaths occurred among males (52.3%; n = 6,608,436) and White decedents (80.1%; n = 10,124,360) followed by Black (11.4%; n = 1,439,259) and Hispanic (5.7%; n = 717,080) decedents. 

## Age of youth during time of parental cancer death

```{r}
sub_age <- results_age_df |>
    filter(race == "total", cause == "cancer", metric == "number", age < 18)

sub_age |>
    uncount(as.integer(median)) |> 
    pull(age) |>
    quantile(probs = c(.25, .5, .75))

```

```{r}
sub_age |>
    group_by(age) |>
    summarize(median = sum(median)) |>
    mutate(
        cume_n = cumsum(median),
        prop = median / sum(median),
        cume_prop = cumsum(prop)
    ) |> 
    kable(row.names = FALSE, 
          digits = c(0, 0, 0, 3, 3),
        format = "html",
        booktabs = TRUE
    ) |>
    kableExtra::kable_styling(bootstrap = c("striped", "scale_down"))
```


## Total number of youth who lost one or more parents due to cancer, 1999 through 2020
```{r}
tot_youth <- results_df |>
    filter(race == "total", cause == "cancer", metric == "number", sex == "both") |>
    mutate(
        cume_median = cumsum(median),
        cume_lower95 = cumsum(lower95),
        cume_upper95 = cumsum(upper95)
    ) |>
    filter(year == max(year)) |>
    mutate(parental_deaths = sprintf(
        "%s (%s to %s)",
        prettyNum(round(cume_median / 1000) * 1000, big.mark = ","),
        prettyNum(round(cume_lower95 / 1000) * 1000, big.mark = ","),
        prettyNum(round(cume_upper95 / 1000) * 1000, big.mark = ",")
    ))

tot_youth |> select(race_cat, death_cat, parental_deaths)
```

## Total number of youth who lost at least one parent from any cause, 1999 through 2020
```{r}
tot_youth_all_causes <- results_df |>
    filter(race == "total", cause == "all", metric == "number", sex == "both") |>
    mutate(
        cume_median = cumsum(median),
        cume_lower95 = cumsum(lower95),
        cume_upper95 = cumsum(upper95)
    ) |>
    filter(year == max(year)) |>
    mutate(parental_deaths = sprintf(
        "%s (%s to %s)",
        prettyNum(round(cume_median / 1000) * 1000, big.mark = ","),
        prettyNum(round(cume_lower95 / 1000) * 1000, big.mark = ","),
        prettyNum(round(cume_upper95 / 1000) * 1000, big.mark = ",")
    ))

tot_youth_all_causes |> select(race_cat, death_cat, parental_deaths)
```

```{r}
tot_youth$cume_median / tot_youth_all_causes$cume_median
```

## Number of parental deaths by year 

```{r}
parental_deaths_by_year <- results_df |>
    filter(race == "total", cause == "cancer", metric == "number", sex == "both") |>
    mutate(parental_deaths = sprintf(
        "%s (%s to %s)",
        prettyNum(round(median / 100) * 100, big.mark = ","),
        prettyNum(round(lower95 / 100) * 100, big.mark = ","),
        prettyNum(round(upper95 / 100) * 100, big.mark = ",")
    ))

parental_deaths_by_year |>
    select(race_cat, death_cat, year, parental_deaths) |>
    kable(row.names = FALSE,
          format = "html",
          booktabs = TRUE) |>
    kableExtra::kable_styling(bootstrap = c("striped", "scale_down"))
```

## Incidence of parental deaths by year

```{r}
incidence_by_year <- results_df |>
    filter(race == "total",
           cause == "cancer",
           metric == "percent",
           sex == "both") |>
    mutate(parental_deaths = sprintf(
        "%0.2f (%0.2f to %0.2f)",
        round(median  * 1000, 2),
        round(lower95 * 1000, 2),
        round(upper95 * 1000, 2)
    ))

incidence_by_year |>
    select(race_cat, death_cat, year, parental_deaths) |>
    kable(row.names = FALSE,
          format = "html",
          booktabs = TRUE) |>
    kableExtra::kable_styling(bootstrap = c("striped", "scale_down"))
```

## Number of parental deaths by year by parent

```{r}
parental_deaths_parent_by_year <- results_df |>
    filter(race == "total", cause == "cancer", metric == "number", sex != "both") |>
    mutate(parental_deaths = sprintf(
        "%s (%s to %s)",
        prettyNum(round(median / 100) * 100, big.mark = ","),
        prettyNum(round(lower95 / 100) * 100, big.mark = ","),
        prettyNum(round(upper95 / 100) * 100, big.mark = ",")
    ))

parental_deaths_parent_by_year |>
    select(race_cat, death_cat, sex, year, parental_deaths) |>
    arrange(year, sex) |> 
    kable(row.names = FALSE,
          format = "html",
          booktabs = TRUE) |>
    kableExtra::kable_styling(bootstrap = c("striped", "scale_down"))
```

## Incidence of parental daths in 2020 by race/ethnicity

```{r}
parental_deaths_parent_by_race <- results_df |>
    filter(year == 2020, cause == "cancer", metric == "percent", sex == "both") |>
    mutate(parental_deaths = sprintf(
        "%0.3f (%0.3f to %0.3f)",
        round(median  * 100, 3),
        round(lower95 * 100, 3),
        round(upper95 * 100, 3)
    ))

parental_deaths_parent_by_race |>
    select(race_cat, death_cat, sex, year, parental_deaths) |>
    arrange(year, sex) |>
    kable(row.names = FALSE,
          format = "html",
          booktabs = TRUE) |>
    kableExtra::kable_styling(bootstrap = c("striped", "scale_down"))
```

## Change from 1999 to 2020 

```{r}
change_df <- results_df |>
    filter(year %in% c(1999, 2020), cause == "cancer", metric == "percent") |>
    group_by(race, sex) |>
    mutate(
        rel_median = 100 * (median - lag(median)) / lag(median),
        rel_upper = 100 * (lower95 - lag(lower95)) / lag(lower95),
        rel_lower = 100 * (upper95 - lag(upper95)) / lag(upper95)
    ) |>
    filter(!is.na(rel_median)) |>
    mutate(sex_cat = factor(
        sex,
        levels = c("both", "father", "mother"),
        labels = c("Both parents", "Father", "Mother"),
        ordered = TRUE
    )) |>
    mutate(parental_deaths = sprintf(
        "%i%% (%i%% to %i%%)",
        round(rel_median, 0),
        round(rel_lower, 0),
        round(rel_upper, 0)
    ))

change_df |>
    select(race_cat, death_cat, sex_cat, parental_deaths) |>
    arrange(race_cat, sex_cat) |>
    kable(row.names = FALSE,
          format = "html",
          booktabs = TRUE) |>
    kableExtra::kable_styling(bootstrap = c("striped", "scale_down"))
```

